---
title: "R Notebook - "
output: html_notebook
editor_options: 
  chunk_output_type: console
---
Kaggle Titanic tutorial 
17.06.2018 12.26
Link: https://www.kaggle.com/mrisdal/exploring-survival-on-the-titanic

- Exploring the Titanic Dataset
- Illustrative data visualisation
- Model to Predict survival on the Titanic (using randomForest)

1. Feature Engineering
2. Missing value imputation
3. Model prediction


Load packages and libraries 
```{r}
# Load libraries

install.packages("mice")

library(ggplot2)            # Visualisation
library(ggthemes)           # Visualisation
library(scales)             # Visualisation
library(dplyr)              # Data manipulation
library(mice)               # Imputation
library(randomForest)       # Classification algorithm

```

Read in data
```{r}
getwd() # Check directory
test <- read.csv('test.csv', stringsAsFactors = F)
train <- read.csv('train.csv', stringsAsFactors = F)

full <- bind_rows(train, test) # bind training and test data

# Check structure of data
str(test)
str (train)
str(full)

```

Variable    Name	Description

Survived	  Survived (1) or died (0)
Pclass	    Passenger’s class
Name	      Passenger’s name
Sex	        Passenger’s sex
Age	        Passenger’s age
SibSp	      Number of siblings/spouses aboard
Parch	      Number of parents/children aboard
Ticket	    Ticket number
Fare	      Fare
Cabin	      Cabin
Embarked	  Port of embarkation


2. Feature Engineering
```{r}
# Extracting passengers names to feed into prediction model
full$Title <- gsub('(.*, )|(\\..*)', '', full$Name)

# Count by title
table(full$Sex, full$Title)
table(full$Sex, full$Survived)
table(full$Survived, full$Sex)
table(full$Survived, full$Pclass)

# surnames of passenger name
full$Surname <- sapply(full$Name,  
                      function(x) strsplit(x, split = '[,.]')[[1]][1])

str(full$Surname)
nlevels(factor(full$Surname))
nlevels(factor(full$Sex)) # confirm 2 genders
nlevels(factor(full$Pclass))

cat(paste('We have', nlevels(factor(full$Surname)), ' unique surnames. I would be interested to infer ethnicity based on surname --- another time.'))

# Crate new family variables e.g. Family size

full$Fsize <- full$SibSp + full$Parch + 1 # create new column 'fsize' to full. inc themselves
nlevels(factor(full$Fsize))

#create a family variable
full$Family <- paste(full$Surname, full$Fsize, sep='_') # concatenate surname and family size with _ seperator

full$Surname
full$Fsize
full$Family

#Visualise the relationship between family size and survival
#dev.new()             #New gfx window to show plot

if (dev.interactive()) dev.new()  #New gfx window to show plot
ggplot(full[1:891,], aes(x = Fsize, fill = factor(Survived))) +
  geom_bar(stat='count', position='dodge') +
  scale_x_continuous(breaks=c(1:11)) +
  labs(x = 'Family Size') +
  theme_few()
```

Survival penalty to singletons and those with family sizes above 4. Collapse this variable into three factors. 
```{r}

# Discretize family size
full$FsizeD[full$Fsize == 1] <- 'singleton'
full$FsizeD[full$Fsize < 5 & full$Fsize > 1] <- 'small'
full$FsizeD[full$Fsize > 4] <- 'large'

# Show family size by survival using a mosaic plot
if (dev.interactive()) dev.new()  #New gfx window to show plot
mosaicplot(table(full$FsizeD, full$Survived), main='Family Size by Survival', shade=TRUE)
table(full$Survived,full$FsizeD)

summary(full$Age)  # We can see 263 NA's / Missing values
table(is.na(full))      # count of Missing data for entire table
table(is.na(full$Age))  # Missing data for specific column
sapply(full, function(x) sum(is.na(x))) # NA  count by columns
rowSums(is.na (full))   # NA count by rows

```
The mosaic plot shows that we preserve our rule that there’s a survival penalty among singletons and large families, but a benefit for passengers in small families. I want to do something further with our age variable, but 263 rows have missing age values, so we will have to wait until after we address missingness.

Check the string length of surnames 
```{r}
# NOTE use ggplot2:: to find what functions in a package ggplot2. useful if you are unsure. Also you don't need to install/load the library if you call 'ggplot2::'

install.packages("stringr")
library(stringr)
summary(full)
tail(full)
#convert to data framea to plot
name_length <- dplyr::data_frame(length = stringr::str_length(full$Surname))   #string length of surnames
table(stringr::str_length(full$Surname), full$Surname)
table(stringlength)
str(name_length)
dev.new()
ggplot(name_length, aes(x= length)) + geom_histogram(bins = 20)

summary(name_length) #get mean value
```
The average surname on the Titanic is around 7 characters long. Perhaps it would be interesting to compare to current average surnames lengths. 

```{r}
# get the first letter of every surname

first_letter <- substr(full$Surname,1,1)
head(full$Surname)
full$Surname[1]
table(first_letter)
#ggplot(first_letter, aes(x= length)) +geom_bar()

# count the first letters 
first_letter_count <- table(first_letter)[table(first_letter)>10]
first_letter_count
str(first_letter_count)

str(first_letter)
sapply(first_letter, class)
ggplot(dplyr::data_frame(letter = first_letter), aes(letter)) + geom_bar()
```

Information about Passenger cabin and their deck

```{r}
full$Cabin[1:28]      #  lots of missing values
strsplit(full$Cabin[2], NULL)[[1]]
# Create a Deck variable. Get passenger deck A - F:
full$Deck<-factor(sapply(full$Cabin, function(x) strsplit(x, NULL)[[1]][1]))
```

3. Missingness
Given the small size of the dataset, we probably should not opt for deleting either entire observations (rows) or variables (columns) containing missing values. We’re left with the option of either replacing missing values with a sensible values given the distribution of the data, e.g., the mean, median or mode. Finally, we could go with prediction. We’ll use both of the two latter methods and I’ll rely on some data visualization to guide our decisions.
```{r}
# Passenger 62 and 830 are missing embarkment
full[c(62, 830), 'Embarked']
table(full$Embarked)
cat(paste('We will infer their values for **embarkment** based on present data that we can imagine may be relevant: **passenger class** and **fare**. We see that they paid<b> $', full[c(62, 830), 'Fare'][[1]][1], '</b>and<b> $', full[c(62, 830), 'Fare'][[1]][2], '</b>respectively and their classes are<b>', full[c(62, 830), 'Pclass'][[1]][1], '</b>and<b>', full[c(62, 830), 'Pclass'][[1]][2], '</b>. So from where did they embark?'))

#Get rid of our missing passenger IDs
embark_fare <- full %>%
  filter(PassengerId != 62 & PassengerId != 830)
str(embark_fare)
factor(embark_fare)

# Use ggplot2 to visualize embarkment, passenger class, & median fare
dev.new()
ggplot(embark_fare, aes(x = Embarked, y = Fare, fill = factor(Pclass))) +
  geom_boxplot() +
  geom_hline(aes(yintercept=80), 
    colour='red', linetype='dashed', lwd=2) +
  scale_y_continuous(labels=dollar_format()) +
  theme_few()
```
The median fare for a first class passenger departing from Charbourg (‘C’) coincides nicely with the $80 paid by our embarkment-deficient passengers. I think we can safely replace the NA values with ‘C’.

```{r}
full$Fare[62]
full$Fare[830]
full$Embarked[62]  # check their embarked
# Since their fare was $80 for 1st class, they most likely embarked from 'C'
full$Embarked[c(62, 830)] <- 'C'

table(is.na(full$Fare))
# we have one more NA fare. Show row 1044
full[1044, ]
full$Fare <- NULL
```
This is a third class passenger who departed from Southampton (‘S’). Let’s visualize Fares among all others sharing their class and embarkment (n = 494).
```{r}
ggplot(full[full$Pclass == '3' & full$Embarked == 'S', ], 
  aes(x = Fare)) +
  geom_density(fill = '#99d6ff', alpha=0.4) + 
  geom_vline(aes(xintercept=median(Fare, na.rm=T)),
    colour='red', linetype='dashed', lwd=1) +
  scale_x_continuous(labels=dollar_format()) +
  theme_few()

# Replace missing fare value with median fare for class/embarkment
full$Fare[1044] <- median(full[full$Pclass == '3' & full$Embarked == 'S', ]$Fare, na.rm = TRUE)
```
3.2 Predictive imputation
```{r}
# Show number of missing Age values
sum(is.na(full$Age))
#We could definitely use rpart (recursive partitioning for regression) to predict missing ages, but I’m going to use the mice package for this task just for something different.

# Make variables factors into factors
factor_vars <- c('PassengerId','Pclass','Sex','Embarked',
                 'Title','Surname','Family','FsizeD')

full[factor_vars] <- lapply(full[factor_vars], function(x) as.factor(x))

# Set a random seed
set.seed(129)

# Perform mice imputation, excluding certain less-than-useful variables:
mice_mod <- mice(full[, !names(full) %in% c('PassengerId','Name','Ticket','Cabin','Family','Surname','Survived')], method='rf') 

# Save the complete output 
mice_output <- complete(mice_mod)

# Plot age distributions
dev.new()
par(mfrow=c(1,2))
hist(full$Age, freq=F, main='Age: Original Data', 
  col='darkgreen', ylim=c(0,0.04))
hist(mice_output$Age, freq=F, main='Age: MICE Output', 
  col='lightgreen', ylim=c(0,0.04))

# Replace Age variable from the mice model.
full$Age <- mice_output$Age

# Show new number of missing Age values
sum(is.na(full$Age))
```
3.3 Feature Engineering
Now that we know everyone’s age, we can create a couple of new age-dependent variables: Child and Mother. A child will simply be someone under 18 years of age and a mother is a passenger who is 1) female, 2) is over 18, 3) has more than 0 children (no kidding!), and 4) does not have the title ‘Miss’.
```{r}
# First we'll look at the relationship between age & survival
dev.new()
ggplot(full[1:891,], aes(Age, fill = factor(Survived))) + 
  geom_histogram() + 
  # I include Sex since we know (a priori) it's a significant predictor
  facet_grid(.~Sex) + 
  theme_few()

# Create the column child, and indicate whether child or adult
full$Child[full$Age < 18] <- 'Child'
full$Child[full$Age >= 18] <- 'Adult'

# Show counts
table(full$Child, full$Survived)

# Adding Mother variable
full$Mother <- 'Not Mother'
full$Mother[full$Sex == 'female' & full$Parch > 0 & full$Age > 18 & full$Title != 'Miss'] <- 'Mother'

# Show counts
table(full$Mother, full$Survived)

# Finish by factorizing our two new factor variables
full$Child  <- factor(full$Child)
full$Mother <- factor(full$Mother)
md.pattern(full)
```
4. Prediction
```{r}
# Split the data back into a train set and a test set
train <- full[1:891,]
test <- full[892:1309,]

# Set a random seed
set.seed(754)

# Build the model (note: not all possible variables are used)
rf_model <- randomForest(factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + 
                                            Fare + Embarked + Title + 
                                            FsizeD + Child + Mother,
                                            data = train)

# Show model error
plot(rf_model, ylim=c(0,0.36))
legend('topright', colnames(rf_model$err.rate), col=1:3, fill=1:3)


# Get importance
importance    <- importance(rf_model)
varImportance <- data.frame(Variables = row.names(importance), 
                            Importance = round(importance[ ,'MeanDecreaseGini'],2))

# Create a rank variable based on importance
rankImportance <- varImportance %>%
  mutate(Rank = paste0('#',dense_rank(desc(Importance))))

# Use ggplot2 to visualize the relative importance of variables
ggplot(rankImportance, aes(x = reorder(Variables, Importance), 
    y = Importance, fill = Importance)) +
  geom_bar(stat='identity') + 
  geom_text(aes(x = Variables, y = 0.5, label = Rank),
    hjust=0, vjust=0.55, size = 4, colour = 'red') +
  labs(x = 'Variables') +
  coord_flip() + 
  theme_few()
```

4.4 Prediction
```{r}
# Predict using the test set
prediction <- predict(rf_model, test)

# Save the solution to a dataframe with two columns: PassengerId and Survived (prediction)
solution <- data.frame(PassengerID = test$PassengerId, Survived = prediction)

# Write the solution to file
write.csv(solution, file = 'rf_mod_Solution.csv', row.names = F)
```



























